Extending Height Above Nearest Drainage to Model Multiple Fluvial Sources in Flood Inundation Mapping Applications for the U.S. National Water Model

### 2023-02-11 manuscripts/owp_fim4_2022/sections/conclusions.tex
second review responses and changes submitted to coauthors 20230211

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
\label{sec:conclusions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Floods present a significant, under-served, and expanding risk to life, property, and resources.
Previous flood forecasting systems lacked the coverage to adequately inform society of these risks.
The National Water Model (NWM), developed by the National Oceanic and Atmospheric Administration's (NOAA) Office of Water Prediction (OWP) and the National Center for Atmospheric Research (NCAR), provides increased spatial coverage and resolution as well as additional forecast time horizons on mostly hourly intervals.
Additional modeling is required to convert streamflows from the NWM to river stages and finally to flood inundation maps (FIM).
Height Above Nearest Drainage (HAND) is a means of detrending digital elevations maps (DEM) by normalizing elevation to the nearest relevant drainage point.
HAND coupled with the use of reach averaged synthetic rating curves (SRC) provide such a means of creating continental scale FIM capabilities at high resolutions (1/3 arc-second, 10 m) and high temporal frequencies (up to 1 hr).
Scalable, open-source software, known as OWP FIM, was developed to produce HAND and associated datasets (catchments, SRCs, and cross-walking data) for the NWM forecasting area including Hawaii and Puerto Rico \cite{inundationMapping2022}.
HAND is produced using the latest hydro-conditioning techniques to enforce monotonically decreasing elevations including stream burning, levee enforcement, pit-filling, stream channel excavation, thalweg breaching, headwater seeding, stream reach resampling, and more. 
Finally, we used this implementation to investigate the skill of the FIMs by varying the scale of the processing units used to derive HAND.
FIM skill was evaluated over large spatial scales by comparison to HEC-RAS 1D models.

The main contribution and conclusion of this work centers around a fundamental limitation in HAND based FIM which is a failure to account for multiple possible sources of fluvial inundation since HAND only considers inundation from the nearest drainage line.
We illustrate that reducing the Horton-Strahler stream order of a HAND processing unit down to one enhances skill by significantly reducing false negatives at junctions of major streams.
In order to reduce stream order of the NWM stream network for HAND computation, we dissected the NWM network into two simpler units of singular Horton-Strahler stream order and mosaiced the resulting FIMs derived from each.
The NWM Mainstems (MS) stream network, which covers roughly 4-5\% of the NWM Full Resolution (FR) network, spans all established forecasting points in the Advanced Hydrologic Prediction System (AHPS) and downstream reaches.
The inundation from MS derived HAND is mosaiced together with the inundation of FR derived HAND.
Extending order reduction to the entire network, the Generalized Mainstem (GMS) technique discretizes the NWM FR network into level paths (LP) of unit stream order for HAND computation.
All LP based FIM derived from LP based HAND datasets are mosaiced together to form one seamless FIM.
Dissecting the stream network into regions of LPs with unit stream order is necessary because HAND has a ``nearest drainage'' limitation meaning it only accounts for riverine inundation sourced from the nearest drainage line.
In our evaluation of this technique, we observe that HAND based FIM improves in skill as we extend from nearest drainage inundation in FR to multiple drainage support in MS for only 4-5\% of the FR network.
Extending multiple drainage support to the entire FR network with GMS based HAND improves skill at around the same magnitude that MS improved upon FR.
Thus we conclude that deriving HAND with independent stream networks of unit Horton-Strahler stream order enhances the skill of FIM but offers diminishing returns as we extend from 4-5\% of the network with MS to 100\% of the network with GMS since deriving HAND and FIMs at these localized scale does add computational expense.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
\label{sec:conclusions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
The main contribution and conclusion of this work centers around a fundamental limitation in HAND based FIM which is a failure to account for multiple possible sources of fluvial inundation since HAND only considers inundation from the nearest drainage line.
We illustrate that reducing the Horton-Strahler stream order of a HAND processing unit down to one enhances skill by significantly reducing false negatives at junctions of major streams.
In order to reduce stream order of the NWM stream network for HAND computation, we dissected the NWM network into two simpler units of singular Horton-Strahler stream order and mosaiced the resulting FIMs derived from each.
The NWM Mainstems (MS) stream network, which covers roughly 4-5\% of the NWM Full Resolution (FR) network, spans all established forecasting points in the Advanced Hydrologic Prediction System (AHPS) and downstream reaches.
The inundation from MS derived HAND is mosaiced together with the inundation of FR derived HAND.
Extending order reduction to the entire network, the Generalized Mainstem (GMS) technique discretizes the NWM FR network into level paths (LP) of unit stream order for HAND computation.
All LP based FIM derived from LP based HAND datasets are mosaiced together to form one seamless FIM.
Dissecting the stream network into regions of LPs with unit stream order is necessary because HAND has a ``nearest drainage'' limitation meaning it only accounts for riverine inundation sourced from the nearest drainage line.
In our evaluation of this technique, we observe that HAND based FIM improves in skill as we extend from nearest drainage inundation in FR to multiple drainage support in MS for only 4-5\% of the FR network.
Extending multiple drainage support to the entire FR network with GMS based HAND improves skill at around the same magnitude that MS improved upon FR.
Thus we conclude that deriving HAND with independent stream networks of unit Horton-Strahler stream order enhances the skill of FIM but offers diminishing returns as we extend from 4-5\% of the network with MS to 100\% of the network with GMS since deriving HAND and FIMs at these localized scale does add computational expense.

### 2023-02-11 manuscripts/owp_fim4_2022/sections/discussion.tex
second review responses and changes submitted to coauthors 20230211

As evident in the results of the study in Section \ref{sec:results}, a sizable amount of the increase in CSI observed by reducing stream order for HAND computation can be attributed to increases in POD.
This can be inferred by close inspection of Figure \ref{fig:violin_plot} and Table \ref{tab:aggregate_metrics} where changes in POD are significantly higher than that of FAR.
%This intuition is confirmed by previously mentioned work by \citeA{gerapetritis2004behavior} where CSI was shown to be a direct function of both POD and FAR where CSI is maximized when $POD = 1 - FAR$.
Upon investigation of the performance of HAND derived FIM, we note a general increase of catchment sizes for the MS and GMS methods when compared to the FR method as they are now delineated independently of any tributaries that would constrain catchment sizes otherwise.
Additionally, we note significantly less water build up along catchment boundaries especially at the junction of lower order tributaries with lower flow rates and higher order rivers with more flow.
This allows for inundation extents to expand across regions previously encapsulated by catchments of joining reaches in lower flow tributaries.
The water built up along the catchment boundaries can be thought of as a column of water in a cylindrical container (catchment) that has exceeded the elevation of the container's rim which does not represent accurate physics.

@@

As an enhancement, this paper proposes computing HAND for stream networks comprised of level-paths independently of one another.
In sub-figure \ref{fig:gms_enhancement}b, the agreement raster for the GMS technique is illustrated as well as the stream network lines in green.
While the entire mosaiced inundation map from GMS (as described in Section \ref{ssec:inundation_mapping} and Equation \ref{eq:hand_fim_depth}) is used to produce this agreement map, we only show the catchments associated with the mainstem of this region that is shown to follow a clear horizontal path.
Showing all the catchments for the tributaries that were all derived independently would convolute the image.
The main message illustrated here is that the catchments associated with the mainstem of this area significantly increase in size.
Since they are not restricted by the catchments of tributaries that lie in the same drainage areas as those of the mainstem, they extend to include those as well.
The consequence for inundation extents is a general increase in spatial coverage of the river's water which shows to be in much better agreement with the benchmark map.
The TPs are no longer bounded by the catchment lines and allowed to expand to their natural extents.

@@

Attempting to diagnose this bias in the SRC leads one to Equation \ref{eq:reach_averaged_mannings_equation} which shows the reach averaged SRC relationship between stage and discharge.
Across the three methods explored, FR, MS, and GMS, one identifies differences in the inputs and outputs and notes no difference in the stages and Manning's n values.
While the channel slope and reach lengths are not exactly the same across methods, their averaged differences are very negligible which only leaves room for deviations in volume and bed area.
Again, volume (V(y) or simply V) is synonymous to reach-averaged cross-sectional area and bed area (B(y) or B) is analogous to reach-averaged hydraulic radius but these associations only hold when reach length, L, is considered.
Discharge, Q, is directly related to volume and inversely related to bed area and each parameter is weighed according to the magnitude of its exponent which are $\frac{5}{3}$ and $\frac{2}{3}$ respectively (see Equation \ref{eq:reach_averaged_mannings_equation}). 
Figures \ref{fig:rating_curve_comparison}b and \ref{fig:rating_curve_comparison}c show how volume and bed area compare across the three methods with GMS having significantly greater values than MS which has greater values than FR.
Again the relative discrepancy between FR vs MS and MS vs GMS is explained by the extent of their spatial coverages.
Both V and B values increase but are weighed differently by their respective exponents and pull Q in different directions.
We show in Figure \ref{fig:rating_curve_comparison}d the relationship of $\frac{V^{5/3}}{B^{2/3}}$ and plot this ratio against stage, y, to show how these two parameters collectively pull the Q up and changes the SRC accordingly.
In other words, the magnitude and weight of the volume at each stage level exceeds the influence of the magnitude and weight of the bed area.
Both parameters are set to increase mainly due to much larger catchments leading to more pixels at each stage level as shown in Figure \ref{fig:rating_curve_comparison}e.

@@

Additional analysis of Figure \ref{fig:gms_enhancement}a, reveals that some catchments don't have inundation or significant inundation.
While the cause of these errors can be varied, we assert here that conflating four networks for use in evaluations leads to significant error.
Section \ref{sssec:cross_walking_networks} details how reach identifiers are conflated for the FIM network back to that of the NWM. 
One of the issues with the FR version of HAND occurs when a reach of given stream order accidentally conflates to that of a neighboring tributary that is of lower order which leads to areas of FNs.
The utilization of MS and GMS only conflates to NWM catchments directly associated with the LP in question which is inherently easy to do with those methods. 
Thus part of the improvement in MS and GMS methods is due to a slight improvement in cross-walking methodology.
The NWM stream network was derived using the NHDPlus V2 dataset which was derived from coarser DEMs than those used here. 
Additional conflation is identified in cross-walking the stream network used by the BLE maps and those of HAND.
Until a singular stream network is used for the NWM, BLE benchmark, and for HAND based FIM, conflation will continue being a source of error.

Our qualitative analysis suggests that the SRCs offer a significant opportunity for improvement in HAND based FIM for future development.
The bathymetry of the 10 m DEM from 3DEP is known to be lacking proper representation thus leading to inadequate representation of volume and bed area with all three methods employed.
Manning's n which typically accounts for roughness could be tuned to account for these DEM limitations or could be held fixed to some local value associated with a given flood magnitude.
Some adjusting parameter must be introduced to enhance the estimation of the bathymetric representation.
Lidar DEMs from the USGS at 3 m and 1 m scale could be utilized to derive HAND as well which we conject should show better agreement with higher fidelity FIMs also derived from the same Lidar based DEMs.
We suspect that a significant amount of the difference in performance between 100 yr and 500 yr magnitude events can be attributed to poor SRC performance due to poor bathymetric representation.
Lower magnitude events are, logically, more susceptible to poor bathymetric data due a greater proportion of the inundation being attributable to areas that are more typically under normal flow conditions.
Higher flow events tend to cover regions with more floodplain inundation thus less sensitive to errors from bathymetric data quality.
On a related note, the use of the AGREE DEM method discussed in Section \ref{ssec:stream_network_enforcment} also interacts with the bathymetry issue introducing several artificial geometry parameters that affect SRC shape and quality.
Due to focus on the nearest drainage problem, we leave future work related to SRC representation including roughness estimation, bathymetric data assimilation, and bathymetry adjustments as opportunities for major enhancements in HAND based FIM.

Lastly, after errors introduced by conflation, poor roughness estimation, bathymetric/elevation adjustment are accounted for, HAND still has another fundamental limitation that is inherently baked into how it works.
For HAND to be derived and thus create a FIM for a given area, that area must entirely drain to the stream network and the stream network must also drain itself.
In other words, an entire area eligible for flooding must monotonically decrease in elevation. 
DEM's naturally don't do this and the dynamics of true flood events don't follow drainage patterns.
Enforcing this assumption for HAND leads to significant amount of DEM manipulations that introduce basic errors.
These errors are deep into the assumptions of HAND and thus more difficult to disentangle.
Ultimately, the use of more advanced 2D hydrodynamic models should be considered for dealing with this limitation of HAND but would come at significant expense at the given high resolution across very large spatial scales and frequent forecast resolutions.
%
\begin{figure}[H]
\centering
\includegraphics[scale=1.0]{figures/rating_curve_comparison.jpg}
\caption{Illustrates average quantities for the three methods, FR, MS, and GMS, for each stage value (m). 
The values are (a) Discharge $m^3s^{-1}$, (b) Volume $m^3$, (c) Bed Area $m^2$, (d) a function of Volume and Bed Area, and (e) number of pixels.
}
\label{fig:rating_curve_comparison}
\end{figure}
%
%
\begin{figure}[H]
\centering
\includegraphics[scale=1.0]{figures/gms_enhancement.jpg}
\caption{OWP FIM inundation agreement, TP, FP, FN, and TN, with BLE HEC-RAS maps in HUC 11140105 at the 500 yr recurrence magnitude.
Catchment boundaries and stream lines are shown in white and dotted green, respectively.
Sub-figure (a) shows agreement of FR HAND denoting significant areas of under-prediction due to junctions and catchment boundaries.
Meanwhile, (b) shows the agreement for GMS and much larger catchments leading to much better inundation agreement for this given reach. 
Overall, this illustrates the benefits of stream order reduction for deriving HAND datasets.
}
\label{fig:gms_enhancement}
\end{figure}
%

>>>>

As evident in the results of the study in Section \ref{sec:results}, a sizable amount of the increase in CSI observed by reducing stream order for HAND computation can be attributed to increases in POD.
This can be inferred by close inspection of Figure \ref{fig:violin_plot} and Table \ref{tab:aggregate_metrics} where changes in POD are significantly higher than that of FAR.
This intuition is confirmed by previously mentioned work by \citeA{gerapetritis2004behavior} where CSI was shown to be a direct function of both POD and FAR where CSI is maximized when $POD = 1 - FAR$.
Upon investigation of the performance of HAND derived FIM, we note a general increase of catchment sizes for the MS and GMS methods when compared to the FR method as they are now delineated independently of any tributaries that would constrain catchment sizes otherwise.
Additionally, we note significantly less water build up along catchment boundaries especially at the junction of lower order tributaries with lower flow rates and higher order rivers with more flow.
This allows for inundation extents to expand across regions previously encapsulated by catchments of joining reaches in lower flow tributaries.
The water built up along the catchment boundaries can be thought of as a column of water in a cylindrical container (catchment) that has exceeded the elevation of the container's rim which does not represent accurate physics.

@@

As an enhancement, this paper proposes computing HAND for stream networks comprised of level-paths independently of one another.
In sub-figure \ref{fig:gms_enhancement}b, the agreement raster for the GMS technique is illustrated as well as the stream network lines in green.
While the entire mosaiced inundation map from GMS (as described in Section \ref{ssec:inundation_mapping} and Equation \ref{eq:hand_fim_depth}) is used to produce this agreement map, we only show the catchments associated with the mainstem of this region that is shown to follow a clear horizontal path.
Showing all the catchments for the tributaries that were all derived independently would convolute the image.
The main message illustrated here is that the catchments associated with the mainstem of this area significantly increase in size.
Since they are not restricted by the catchments of tributaries that lie in the same drainage areas as those of the mainstem, they extend to include those as well.
The consequence for inundation extents is a general increase in spatial coverage of the river's water which shows to be in much better agreement with the benchmark map.
The TPs are no longer bounded by the catchment lines and allowed to expand to their natural extents.
Additionally, this study has a limitation in that it did not assess FIM depths, so limited conclusions can be made about the impact of drainage order reduction techniques on the accuracy of FIM depths or water surface elevations.

@@

Attempting to diagnose this bias in the SRC leads one to Equation \ref{eq:reach_averaged_mannings_equation} which shows the reach averaged SRC relationship between stage and discharge.
Across the three methods explored, FR, MS, and GMS, one identifies differences in the inputs and outputs and notes no difference in the stages and Manning's n values.
While the channel slope and reach lengths are not exactly the same across methods, their averaged differences were found to be statistically insignificant which only leaves room for deviations in volume and bed area.
Specifically, we found FR to have an average reach length for the study region of 1354.8 km with a standard deviation 249.6km.
For GMS, we computed an average reach length of 1398.1 km with a standard deviation of 50.2 km.
Again, volume (V(y) or simply V) is synonymous to reach-averaged cross-sectional area and bed area (B(y) or B) is analogous to reach-averaged hydraulic radius but these associations only hold when reach length, L, is considered.
Discharge, Q, is directly related to volume and inversely related to bed area and each parameter is weighed according to the magnitude of its exponent which are $\frac{5}{3}$ and $\frac{2}{3}$ respectively (see Equation \ref{eq:reach_averaged_mannings_equation}). 
Figures \ref{fig:rating_curve_comparison}b and \ref{fig:rating_curve_comparison}c show how volume and bed area compare across the three methods with GMS having significantly greater values than MS which has greater values than FR.
Again the relative discrepancy between FR vs MS and MS vs GMS is explained by the extent of their spatial coverages.
Both V and B values increase but are weighed differently by their respective exponents and pull Q in different directions.
We show in Figure \ref{fig:rating_curve_comparison}d the relationship of $\frac{V^{5/3}}{B^{2/3}}$ and plot this ratio against stage, y, to show how these two parameters collectively pull the Q up and changes the SRC accordingly.
In other words, the magnitude and weight of the volume at each stage level exceeds the influence of the magnitude and weight of the bed area.
Both parameters are set to increase mainly due to much larger catchments leading to more pixels at each stage level as shown in Figure \ref{fig:rating_curve_comparison}e.

@@

Additional analysis of Figure \ref{fig:gms_enhancement}a, reveals that some catchments don't have inundation or significant inundation.
While the cause of these errors can be varied, we assert here that conflating four networks for use in evaluations leads to significant error.
Section \ref{sssec:cross_walking_networks} details how reach identifiers are conflated for the FIM network back to that of the NWM. 
One of the issues with the FR version of HAND occurs when a reach of given stream order accidentally conflates to that of a neighboring tributary that is of lower order which leads to areas of FNs.
The utilization of MS and GMS only conflates to NWM catchments directly associated with the LP in question which is inherently easy to do with those methods. 
Thus part of the improvement in MS and GMS methods is due to a slight improvement in cross-walking methodology.
The NWM stream network was derived using the NHDPlus V2 dataset which was derived from coarser DEMs than those used here. 
Additional conflation errors are realized by cross-walking the stream network used by the BLE maps and those of HAND.
The methods that intersect HEC-RAS cross-sections with NWM reaches could introduce errors that violate mass conservation.
Additionally, as previously noted and illustrated in Figure \ref{fig:ble_evaluation_method}, the BLE network is denser than that of the NWM which leads to FNs due to headwater streams that are present in the BLE but not within the NWM.
Until a singular stream network is used for the NWM, BLE benchmark, and for HAND based FIM, conflation will continue being a source of error.

Our qualitative analysis suggests that the SRCs offer a significant opportunity for improvement in HAND based FIM for future development.
The bathymetry of the 10 m DEM from 3DEP is known to be lacking proper representation thus leading to inadequate representation of volume and bed area with all three methods employed.
Manning's n which typically accounts for roughness could be tuned to account for these DEM limitations or could be held fixed to some local value associated with a given flood magnitude.
Some adjusting parameter must be introduced to enhance the estimation of the bathymetric representation.
Lidar DEMs from the USGS at 3 m and 1 m scale could be utilized to derive HAND as well which we conject should show better agreement with higher fidelity FIMs also derived from the same Lidar based DEMs.
We suspect that a significant amount of the difference in performance between 100 yr and 500 yr magnitude events can be attributed to poor SRC performance due to poor bathymetric representation.
Lower magnitude events are, logically, more susceptible to poor bathymetric data due a greater proportion of the inundation being attributable to areas that are more typically under normal flow conditions.
Higher flow events tend to cover regions with more floodplain inundation thus less sensitive to errors from bathymetric data quality.
On a related note, the use of the AGREE DEM method discussed in Section \ref{ssec:stream_network_enforcement} and \ref{sec:app_stream_network_enforcement} also interacts with the bathymetry issue introducing several artificial geometry parameters that affect SRC shape and quality.
To reiterate in this discussion, Section \ref{ssec:evaluation} pointed out that the BLE's representation of reservoir inundation extends beyond the NWM's defined reservoir boundaries, which results in an under-prediction in the OWP FIM that is reflected in the computed metrics.
Due to focus on the nearest drainage problem, we leave future work related to SRC representation including roughness estimation, bathymetric data assimilation and adjustments as well as reservoir inundation as opportunities for major enhancements in HAND based FIM.

Lastly, after errors introduced by conflation, poor roughness estimation, bathymetric/elevation adjustment, and reservoir representation are accounted for, HAND still has another fundamental limitation that is inherently baked into how it works.
For HAND to be derived and thus create a FIM for a given area, that area must entirely drain to the stream network and the stream network must also drain itself.
In other words, an entire area eligible for flooding must monotonically decrease in elevation. 
DEM's naturally don't do this and the dynamics of true flood events don't follow drainage patterns.
Enforcing this assumption for HAND leads to significant amount of DEM manipulations that introduce basic errors.
These errors are deep into the assumptions of HAND and thus more difficult to disentangle.
Ultimately, the use of more advanced 2D hydrodynamic models should be considered for dealing with this limitation of HAND but would come at significant expense at the given high resolution across very large spatial scales and frequent forecast resolutions.
%
\begin{figure}[H]
\centering
\includegraphics[scale=1.0]{figures/gms_enhancement.jpg}
\caption{OWP FIM inundation agreement, TP, FP, FN, and TN, with BLE HEC-RAS maps in HUC 11140105 at the 500 yr recurrence magnitude.
Catchment boundaries and stream lines are shown in white and dotted green, respectively.
Sub-figure (a) shows agreement of FR HAND denoting significant areas of under-prediction due to junctions and catchment boundaries.
Meanwhile, (b) shows the agreement for GMS and much larger catchments leading to much better inundation agreement for this given reach. 
Overall, this illustrates the benefits of stream order reduction for deriving HAND datasets.
}
\label{fig:gms_enhancement}
\end{figure}
%
%
\begin{figure}[H]
    \centering
    \includegraphics[scale=1.0]{figures/rating_curve_comparison.jpg}
    \caption{Illustrates average quantities for the three methods, FR, MS, and GMS, for each stage value (m). 
    The values are (a) Discharge $m^3s^{-1}$, (b) Volume $m^3$, (c) Bed Area $m^2$, (d) a function of Volume and Bed Area, and (e) number of pixels.
    }
    \label{fig:rating_curve_comparison}
    \end{figure}
    %

### 2023-02-11 manuscripts/owp_fim4_2022/sections/introduction.tex
second review responses and changes submitted to coauthors 20230211

Related to these efforts, the United States Geological Survey (USGS) has invested in relative elevation HAND-like methods via work in the GIS Flood Tool (GFT) that also uses SRCs with cross-sections for stage-discharge relationships \cite{verdin2016software}.
Additional investment by \citeA{petrochenkov2020pygft} was able to successfully scale this approach by transitioning the method to open-source Python source code (PyGFT) and implementing novel interpolation methods to help address some of the catchment boundary discontinuities discussed more in this paper.
In addition to the domestic work done in the US, some studies have expanded upon HAND to cover global domains at 30 m resolutions \cite{yamazaki2019merit,donchyts2016global}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Office of Water Prediction Flood Inundation Mapping}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
In order to mitigate the ever increasing threat of flooding to life and property, an operational capability is required to extend NWM streamflow forecasts to river stages, inundation extents, and inundation depths.
OWP FIM is introduced here as a continental scale capability that generates these products at high spatial and temporal resolutions.
Here we introduce OWP FIM that utilizes a few of the latest techniques in HAND based FIM oriented for use with the NWM in continental scale operational forecasting settings. 
Within the operational framework of OWP FIM, we introduce research demonstrating how FIM performance skill with HAND can be improved by discretizing stream networks into units of an effective unit Horton-Strahler stream order \cite{horton1945erosional,strahler1952hypsometric,strahler1952hypsometric} for HAND computation contexts.
Previous authors dating back to the first HAND for FIM work by \citeA{nobre2016hand} have noted a sensitivity of mapping skill to the stream accumulation threshold which is closely related to stream density and the maximum Horton-Strahler stream order (or simply stream order) of the processing unit employed \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation}.
Here we demonstrate how reducing a HAND processing unit's stream network into discrete level paths (LPs) of singular, effective stream order, can enhance FIM skill by accounting for multiple possible sources of fluvial inundation.
This capability is introduced progressively as MS (whose network represents about 4\% of FR network) and to a higher degree Generalized Mainstems (GMS) (covers entire FR network) which will be explained later on.
The following methods and results describe the work in more detail and demonstrate its efficacy in producing enhanced FIM for the NWM.
%

>>>>

Related to these efforts, the United States Geological Survey (USGS) has invested in relative elevation HAND-like methods via work in the GIS Flood Tool (GFT) that also uses SRCs with cross-sections for stage-discharge relationships \cite{verdin2016software}.
Additional investment by \citeA{petrochenkov2020pygft} was able to successfully scale this approach by transitioning the method to open-source Python source code (PyGFT) and implementing novel interpolation methods to help address some of the catchment boundary discontinuities discussed more in this paper.
In addition to the domestic work done in the US, some studies have expanded upon HAND to cover global domains at 30 m resolutions \cite{yamazaki2019merit,donchyts2016global}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Office of Water Prediction Flood Inundation Mapping}
\label{ssec:owp_fim}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
In order to mitigate the ever increasing threat of flooding to life and property, an operational capability is required to extend NWM streamflow forecasts to river stages and inundation extents.
OWP FIM is introduced here as a continental scale capability that generates these products at high spatial and temporal resolutions.
Here we introduce OWP FIM that utilizes a few of the latest techniques in HAND based FIM oriented for use with the NWM in continental scale operational forecasting settings. 
Within the operational framework of OWP FIM, we introduce research demonstrating how FIM performance skill with HAND can be improved by discretizing stream networks into units of an effective unit Horton-Strahler stream order \cite{horton1945erosional,strahler1952hypsometric,strahler1952hypsometric} for HAND computation contexts.

Previous authors dating back to the first HAND for FIM work by \citeA{nobre2016hand} have noted a sensitivity of mapping skill to the stream accumulation threshold which is closely related to stream density and the maximum Horton-Strahler stream order (or simply stream order) of the processing unit employed \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation}.
HAND produced FIMs are limited in only providing inundation sourced from the nearest drainage line, however, depending on flow conditions and topography, a given area may have multiple contributing fluvial sources of surface water inundation.
The forecasting community, in reviewing HAND, have noted significant negative effects at the confluence of lower flow tributaries with higher flow rivers for which the phrase ``catchment boundary issue'' has been termed.
In previous studies, FIM skill has been shown to be sensitive to the drainage density of the stream network employed as the datum for HAND which is closely related to the maximum Horton-Strahler stream order of the network \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation,nobre2016hand}.
This sensitivity is partly in due to the limitation that catchment boundaries place on inundation extents where only the nearest drainage line can source inundation for any particular area.

Figure \ref{fig:catchment_boundaries_issue} illustrates the exact situation our solution proposes to address where two tributaries converge with a higher order stream segment. 
An actual map with OWP FIM is generated using the NWM full-resolution stream network and compared with a FEMA 100 yr extent (see Section \ref{ssec:evaluation} for more details) showing significant under-prediction in inundation extent.
The higher discharge along the main segment in Figure \ref{fig:catchment_boundaries_issue} of 1,900 cubic meters per second (CMS) does not translate to the lower flow rates along the tributaries of 84 and 195 CMS. 
This is due to a lack of representation of backwater conditions in the hydraulic routing techniques used.
As a parallel problem, there is excess water accumulated along the mainstem that cannot extend in either a fluvial or pluvial manner beyond the boundaries of the mainstem catchments.
%
\begin{figure}[H]
\centering
\includegraphics[scale=1.0]{figures/catchment_boundaries_issue.jpg}
\caption{The figure represents an agreement map between a HAND derived FIM and one produced from the Base Level Engineering (BLE) program for a 100 yr magnitude event at HUC8 12090301.
Agreement maps are symbolized by false negatives (FN), true negatives (TN), false positives (FP), and true positives (TP) where inundated represents the positive condition (see Section \ref{ssec:evaluation} for more details).
The streamflows associated with each river segment are shown in CMS while the flow directions are symbolized as red arrows.
The presence of FNs at the confluence of tributaries (circled in red) with the main segment is associated with lower flow rates in the tributaries that don't account for backwater effects.
Additionally, the flow of 1900 CMS from the main segment cannot extend to the neighboring catchments belonging to its tributaries shown here.
Water pools up vertically along the catchment boundaries of the higher order segment distorting rating curve behavior (Section \ref{sec:discussion}).
Sourcing fluvial inundation from HAND is limited to only its nearest drainage line which is the main issue this study aims to address.
}
\label{fig:catchment_boundaries_issue}
\end{figure}
%

We seek to resolve this catchment boundary problem or nearest drainage limitation by discretizing the target stream network into stream networks of reduced, unit stream order to avoid the constraining of catchments by those belonging to lower order neighbors.
By discretizing the network into stream networks of unit stream order (later defined as level paths or LPs), we remove the influence of neighboring catchments that constrain the inundation extent.
This creates much larger and overlapping catchments that can source fluvial inundation from multiple reaches as required by the given river stage at current flow conditions.
We present two successive methods, National Weather Service MS (Section \ref{sssec:nws_mainstems}) and Generalized Mainstems (GMS) (Section \ref{sssec:generalized_mainstems}), implemented that reduce the effective Horton-Strahler stream orders of the networks employed and test our presented hypothesis that stream networks of unit stream order enhance quality of FIM extents produced with HAND by expanding the nearest drainage definition to increase potential inundated areas.

Here we demonstrate how reducing a HAND processing unit's stream network into discrete LPs of singular, effective stream order, can enhance FIM skill by accounting for multiple possible sources of fluvial inundation.
This capability is introduced progressively as MS (whose network represents about 4\% of FR network) and to a higher degree GMS (covers entire FR network) which will be explained later on.
The following methods and results describe the work in more detail and demonstrate its efficacy in producing enhanced FIM for the NWM.
%
